Psychological Needs During Intergroup Contact

### 2021-11-22 manuscript.tex
introduction structure added
first draft Intergroup Contact Mechanisms

In its most essential interpretation, the intergroup contact hypothesis postulates that frequent and positive contact with an out-group reduces prejudice and increases favorable attitudes towards the other group \citep[e.g.,][]{Hewstone1996, Pettigrew1998}. A key condition for these contact benefits, has been that the interaction is indeed perceived as positive --- making the interaction quality a crucial mechanism of inter-group contact \citep[e.g.,][]{MacInnis2015}. It is widely accepted that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for a positive contact \citep[Allport's Optimal Contact conditions,][]{Allport1954b}. And indeed a major meta-analytic review showed, that intergroup contact benefits were larger when Allport's conditions were met. However, the meta-analysis also showed that contact resulted in more positive intergroup relations even when Allport's conditions were not met. It, thus, remains unclear when and why exactly an interaction is perceived as positive and what factors (beyond Allport's conditions) explain positive contact effects.

@@

\section{Intergroup Contact Mechanisms}
\begin{itemize}
    \item theoretical mechanisms (moderator and mediators) \citep[][]{Brown2005}. They don't discuss Allport's conditions(?)
    \item Allport's conditions
    \item interaction quality
\end{itemize}

\section{Psychological Needs in Intergroup Relations}
\begin{itemize}
    \item general \citep[][]{Shnabel2008}. + Nurit's recent papers 
    \item migrants \citep[][]{Celebi2017}
    \item intergroup contact \citep[][]{Dovidio2017}
\end{itemize}

@@

\section{Process Data}
\begin{itemize}
    \item retrospective vs. near-time data
    \item need for real-life data
    \item barriers
    \item innovations like 'contact logger' \citep[][]{Keil2020}, and FormR \citep[][]{Arslan2020} make this possible now
\end{itemize}

>>>>

In its most essential interpretation, the intergroup contact hypothesis postulates that frequent and positive contact with an out-group reduces prejudice and increases favorable attitudes towards the other group \citep[e.g.,][]{Hewstone1996, Pettigrew1998}. A key condition for these contact benefits, has been that the interaction is indeed perceived as positive --- making the interaction quality a crucial mechanism of inter-group contact \citep[e.g.,][]{MacInnis2015}. It is widely accepted that equal status, common goals, collaboration, and structural support during the interaction form the optimal conditions for a positive contact \citep[Allport's Optimal Contact conditions,][]{Allport1954b, Pettigrew1969}. And indeed a major meta-analytic review showed, that intergroup contact benefits were larger when Allport's conditions were met. However, the meta-analysis also showed that contact resulted in more positive intergroup relations even when Allport's conditions were not met \citep[][]{Pettigrew2006}. It, thus, remains unclear when and why exactly an interaction is perceived as positive and what factors (beyond Allport's conditions) explain positive contact effects.

@@

\section{Intergroup Contact Mechanisms}
\faQuestionCircle\ Not sure whether we need this section

As one of the most researched and practically implemented theories of social psychology, intergroup contact has received a vast amount of empirical, and theoretical attention \citep[][]{Dovidio2017}. Several meta-analytic reviews have found moderately sized positive effects of intergroup contact \citep[][]{Tropp2005, Pettigrew2006, Davies2011}. Additional meta-analyses have additionally shown that intergroup contact is also effective outside the lab \citep[][]{Beelmann2014, Lemmer2015}, during online contact \citep[][]{White2020}, as well as during several more indirect forms of contact \citep[][]{Miles2014, Zhou2019, Harwood2021}.

With such a persistent and meaningful effect, researchers have long proposed theoretical models to understand these intergroup contact effects. These searches for moderators and mediators have, for example, highlighted different forms of social categorizations \citep[][]{Pettigrew1998}, the salience of social categories \citep[][]{Brown2005}, intimacy \citep[e.g.,][]{Marinucci2021} and attachment \citep[e.g.,][]{Tropp2021}, threat and intergroup anxiety \citep[e.g.,][]{Stephan2008, Paolini2004}, and to a lesser extent knowledge about the other group \citep[][]{Pettigrew2008c}.

Interestingly, one concept that has arguably remained at the center of theoretical debates throughout the decades has been the role of 'positive' contact. \st{Although Gordon Allport was not generally optimistic of intergroup contact (especially if it was casual or by too deeply prejudiced individuals), he did see immense potential if only such interactions led to positive associations.} In his now famous chapter on 'the effects of contact', Gordon Allport (\citeyear{Allport1954b}) set out to summarize under what circumstances an interaction between groups tends to be positive, leading to favourable attitudes and reduced prejudice. In the years that followed, several additional conditions were proposed under which a positive intergroup contact would occur \citep[for a critical review see][]{Pettigrew1986}. As a counter-movement to this rather restrictive approach, with the beginning of the 21\textsuperscript{st} century attention had shifted towards investigating the psychological mechanisms of intergroup contact \citep[e.g. see,][]{Paolini2021}. With this shift in focus, the idea of 'positive contact' has seemingly seen a renaissance, where the adjective positive is now not merely an evaluation of the outcome, but rather a feature of the interaction itself. As such, recent work has focused on the role of 'perceived interaction quality' \citep[e.g.,][]{Brown2007, Voci2003} or 'interaction valence' \citep[e.g.,][]{Tropp2016, Barlow2012}, to show that whether a

MAYBE: Yet, relatively little research has investigated why interactions are perceived as positive. That is to say that although a sizeable number of moderators and mediators of intergroup contact have been proposed and interaction quality is seen as an important to the betterment of intergroup relations,

ALTERNATIVE: The literature on intergroup contacts, has thus far mostly ignored the role of individual and social motivations within the intergroup contact process.

\section{Psychological Needs in Intergroup Relations}
\begin{itemize}
    \item general \citep[][]{Shnabel2008, Downie2008}. + Nurit's recent papers 
    \item migrants \citep[][]{Celebi2017}
    \item intergroup contact \citep[][]{Dovidio2017}
\end{itemize}

@@

\section{Process Data}
\begin{itemize}
    \item retrospective vs. near-time data
    \item need for longitudinal \citep[][]{Pettigrew1998, Pettigrew2008, Pettigrew2008b, Pettigrew2011} and real-life data \citep[][]{MacInnis2015, McKeown2017}
    \item barriers
    \item innovations like 'contact logger' \citep[][]{Keil2020}, and FormR \citep[][]{Arslan2020} make this possible now \citep[also see, e.g.,][]{ODonnell2021}
\end{itemize}
